Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated

### 2024-12-06 submission/R2/manuscript.qmd
submission
R3 prep

mask: false

@@

**Advancements in technology have now made it possible to create near-perfect simulations that are indistinguishable from reality with an ease, affordability and accessibility that are unprecedented in Human history**.These artificial, yet realistic constructs permeate all areas of
life through immersive works of fiction, deep fakes (real-like images
and videos generated by deep learning algorithms), virtual and augmented
reality (VR and AR), artificial beings (artificial intelligence "bots"
with or without a physical form), fake news and skewed narratives, of
which ground truth is often hard to access [@nightingale2022]. Such
developments not only carry important consequences for the technological
and entertainment sectors, but also for security and politics - for
instance if used for propaganda and disinformation, recruitment into
malevolent organizations, or religious indoctrination [@pantserev2020].
This issue is central to what has been coined the "post-truth era"
[@lewandowsky2017beyond], in which the distinction (and lack thereof)
between authentic and simulated objects will play a critical role.

@@

Literature shows that the context surrounding a stimulfus often plays an
important role in the assessment of its reality [a process henceforth
referred to as *simulation monitoring*, @makowski2018cognitive;
@makowski2019phenomenal]. With the extensive search and processing of
cues within ambiguous stimuli being an increasingly complex and
cognitively effortful strategy [@michael2021source;
@susmann2021persuasion], people tend to draw on peripheral contextual
cues (**Figure 1**), such as the source of the stimulus (e.g., which
journal was the information published in), and its credibility,
authority and expertise, to help facilitate their evaluation
[@petty1986elaboration; @susmann2021persuasion; @michael2021source].
However, the atomization and decontextualization of information allowed
by online social media (where text snippets or video excerpts are often
mass-shared with little context) makes this task progressively difficult
[@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear
contextual information, what drives our beliefs of reality?

@@

**Ethics Statement.** This study was approved by the NTU Institutional
Review Board (NTU IRB-2022-187) <!-- [**IRB NUMBER**] --> and all
procedures performed were in accordance with the ethical standards of
the institutional board and with the 1964 Helsinki Declaration. All
participants provided their informed consent prior to participation and
were incentivized after completing the study.

**Procedure.** In the first part of the study, participants answered a
series of personality questionnaires presented in the order below. These
include the *Mini-IPIP6* <!-- (24 items) [@sibley2011]  --> [24 items,
@sibley2011] measuring 6 personality traits, the *SIAS-6* and the
*SPS-6* <!-- (6 items each) [@peters2012development] --> [6 items each,
@peters2012development] assessing social anxiety levels, 5 items we
devised pertaining to expectations about AI-generated image technology
("I think current Artificial Intelligence algorithms can generate very
realistic images"), **to potentially test and mitigate the potential
effect of expectations/beliefs about AI.** **These items were** mixed
with 5 items from the general attitudes towards AI scale to lower the
former's saliency and the possibility of it priming the subjects about
the task, <!-- (*GAAIS*) [@schepman2020initial]. --> [*GAAIS*,
@schepman2020initial] the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] --> [30 items,
@jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] --> [18 items,
@freeman2021revised] measuring 2 dimensions related to paranoid
thinking; and the *IUS-12* <!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing] measuring intolerance to uncertainty.
Self-rated attractiveness was also assessed using 2 items - one
measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->
["How attractive would you say you are?", @marcinkowska2021self] and the
other measuring physical attractiveness
<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the
average", @spielmann2020predictive]. 3 attention check questions were
also embedded in the surveys. **All Cronbach's alpha values were within
the acceptable to excellent range, except for the neuroticism subscale
of the Mini-IPIP6 and the negative subscale of the GAAI, which were
poor, and the Expectations about AI scale, which was questionable (Gliem
& Gliem, 2003; see supplementary material for the details of the
reliability analysis).**

In the second part of this study, images of neutral-expression faces
from the validated American Multiracial Face Database
<!-- (AMFD) [@chen2021broadening]  --> [AMFD, @chen2021broadening] were
presented to the participants for 500ms each, in a randomized order,
following a fixation cross display (750 ms). T**he decision to present
the faces for 500 ms was based on pilot studies, which demonstrated that
this duration provides a sufficient perceptual window for
decision-making and aligns with previous research indicating stable
judgment levels and increased confidence beyond this exposure time
[@willis2006first].**

@@

**Participants.** Although the main part of the study relied on
within-subject design (with 109 trials per participant), we also planned
to do between-participants analyses, thus aiming at collecting a larger
sample than traditionally used in experimental psychology (with budget
availability as the main constraint). One hundred and fifty participants
were recruited via *Prolific*, a crowd-sourcing platform recognized for
providing high quality data [@peer2022; douglas2023data]. The only
inclusion criterion was a fluent proficiency in English to ensure that
the experiment instructions would be well-understood. Participants were
incentivised with a reward of about \textsterling 7.5 for completing the
study, which took about 45 minutes to finish. Demographic variables
(age, gender, sexual orientation, education and ethnicity) were
self-reported on a voluntary basis.

@@

The determinants of reality beliefs were modelled separately for
attractiveness, beauty, trustworthiness, and familiarity, using second
order raw polynomials coefficients to allow for possible quadratic
relationships (**Figure 2**). Aside from attractiveness (conceptualized
as a general construct), models for beauty, trustworthiness and
familiarity were adjusted for the the two remaining variables *mutatis
mutandis*. The analysis focused on sexual-orientation relevant stimuli,
i.e., on faces that were aligned with respect to the participants'
sexual orientation (i.e., female faces for heterosexual males, male
faces for homosexual males, etc.), and the models included the
interaction with the participants' gender (as a sexual dimorphism has
been reported in face appraisal processes). For the attractiveness and
beauty models, we then added the interaction with the reported
self-attractiveness (the average of the two questions pertaining to it)
to investigate its potential modulatory effect. Finally, we investigated
the inter-individual correlates of simulation monitoring with similar
models (but this time, for all items regardless of the participant's
gender or sexual orientation) for each questionnaire, with all of the
subscales as orthogonal predictors.\

```{r message=FALSE, warning=FALSE, fig.cap="Figure 2. Top part shows the efffect of face ratings on 1) the probability of judging a face as real vs. fake (solid line) and 2) on the confidence associated with that judgement (dashed lines) depending on the sex. Bottom part shows the effect of personality traits on the belief (black line) and the confidence associated with it (colored lines). The points are the average per participant confidence for both types of judgements. Stars indicate significance (p < .001***, p < .01**, p < .05*).", out.width="100%"}
knitr::include_graphics("../../figures/Figure2.png")
```

@@

Interestingly, we found a significant positive linear relationship in
male participants for both attractiveness and beauty on simulation
monitoring that we could interpret under an evolutionary lens.
Specifically, males purportedly place more emphasis on facial
attractiveness as a sign of reproductive potential, as compared with
females, who tend to value characteristics signaling resource
acquisition capabilities [@qi2022gender; @fink2006facial;
@buunk2002age]. It is thus possible that the evolutionary weight
associated with attractiveness skewed the perceived saliency of
attractive faces for men, rendering them significantly more salient than
unattractive faces, and in turn distorting the relationship with
simulation monitoring.
<!-- However, it is also possible that this pattern occurred due to the stimuli range, by which the stimuli did not include sufficiently unattractive faces for males to be , by which   -->
However, future studies should test this saliency-based hypothesis by
measuring constructs closer to salience and its effects, for instance
using neuroimaging [@lou2015perceptual; @indovina2007dissociation] or
physiological markers
<!-- (e.g., heart rate deceleration) [@skora2022functional]. --> [e.g.,
heart rate deceleration, @skora2022functional].

@@

Another issue is the impact on reality judgements of the prior explicit
instruction that "about half of the faces were AI-generated and the
other half real photos". Given this prior information given to
participants, it might seems like our enthusiasm pertaining to the
finding that most people did indeed believe a high number of stimuli to
be fake might be unwarranted, since it simply affirms participants
followed the instructions. However, even if that was the case, the
finding that our beliefs of reality can be so easily re-programmed with
simple instructions and lead to high-confidence answers remains an
interesting phenomenon. **Moreover, it is to note that the paradigm did not explicitly instruct participants to balance their answers according to a certain distribution (e.g., 50-50) - merely providing them a description of the dataset (but participants could, and in some cases did, deviate substantially from the information provided).** The fact that no presentation order effect was found on
reality beliefs suggests that participants did not try to actively
distribute their responses to match the instructions, in which case we
would have expected a different pattern: for instance, the first few
items judged as real (the initial "true" belief of the participants),
and a bias would progressively appear towards responding "fake" (as
participants realize that all stimuli are of similar nature and that
they have to "make up" for the prevalence of their "real" answers to
fulfil the expected proportion of responses given the instructions).

@@

# Data Availability

@@

@@

# References

>>>>

mask: true

@@

**Advancements in technology have now made it possible to create
near-perfect simulations that are indistinguishable from reality with an
ease, affordability and accessibility that are unprecedented in Human
history**. These artificial, yet realistic constructs permeate all areas
of life through immersive works of fiction, deep fakes (real-like images
and videos generated by deep learning algorithms), virtual and augmented
reality (VR and AR), artificial beings (artificial intelligence "bots"
with or without a physical form), fake news and skewed narratives, of
which ground truth is often hard to access [@nightingale2022]. Such
developments not only carry important consequences for the technological
and entertainment sectors, but also for security and politics - for
instance if used for propaganda and disinformation, recruitment into
malevolent organizations, or religious indoctrination [@pantserev2020].
This issue is central to what has been coined the "post-truth era"
[@lewandowsky2017beyond], in which the distinction (and lack thereof)
between authentic and simulated objects will play a critical role.

@@

Literature shows that the context surrounding a stimulus often plays an
important role in the assessment of its reality [a process henceforth
referred to as *simulation monitoring*, @makowski2018cognitive;
@makowski2019phenomenal]. With the extensive search and processing of
cues within ambiguous stimuli being an increasingly complex and
cognitively effortful strategy [@michael2021source;
@susmann2021persuasion], people tend to draw on peripheral contextual
cues (**Figure 1**), such as the source of the stimulus (e.g., which
journal was the information published in), and its credibility,
authority and expertise, to help facilitate their evaluation
[@petty1986elaboration; @susmann2021persuasion; @michael2021source].
However, the atomization and decontextualization of information allowed
by online social media (where text snippets or video excerpts are often
mass-shared with little context) makes this task progressively difficult
[@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear
contextual information, what drives our beliefs of reality?

@@

**Ethics Statement.** This study was approved by the NTU Institutional
Review Board (NTU IRB-2022-187) <!-- [**IRB NUMBER**] -->and all
procedures performed were in accordance with the ethical standards of
the institutional board and with the 1964 Helsinki Declaration. All
participants provided their informed consent prior to participation and
were incentivized after completing the study.

**Procedure.** In the first part of the study, participants answered a
series of personality questionnaires presented in the order below. These
include the *Mini-IPIP6* <!-- (24 items) [@sibley2011]  --> [24 items,
@sibley2011] measuring 6 personality traits, the *SIAS-6* and the
*SPS-6* <!-- (6 items each) [@peters2012development] --> [6 items each,
@peters2012development] assessing social anxiety levels, 5 items we
devised pertaining to expectations about AI-generated image technology
("I think current Artificial Intelligence algorithms can generate very
realistic images"), **to potentially test and mitigate the potential
effect of expectations/beliefs about AI.** **These items were** mixed
with 5 items from the general attitudes towards AI scale to lower the
former's saliency and the possibility of it priming the subjects about
the task, <!-- (*GAAIS*) [@schepman2020initial]. --> [*GAAIS*,
@schepman2020initial] the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] --> [30 items,
@jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] --> [18 items,
@freeman2021revised] measuring 2 dimensions related to paranoid
thinking; and the *IUS-12* <!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing] measuring intolerance to uncertainty.
Self-rated attractiveness was also assessed using 2 items - one
measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->["How
attractive would you say you are?", @marcinkowska2021self] and the other
measuring physical
attractiveness<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the
average", @spielmann2020predictive]. 3 attention check questions were
also embedded in the surveys. **All Cronbach's alpha values were within
the acceptable to excellent range, except for the neuroticism subscale
of the Mini-IPIP6 and the negative subscale of the GAAI, which were
poor, and the Expectations about AI scale, which was questionable (Gliem
& Gliem, 2003; see supplementary material for the details of the
reliability analysis).**

In the second part of this study, images of neutral-expression faces
from the validated American Multiracial Face Database
<!-- (AMFD) [@chen2021broadening]  -->[AMFD, @chen2021broadening] were
presented to the participants for 500ms each, in a randomized order,
following a fixation cross display (750 ms). T**he decision to present
the faces for 500 ms was based on pilot studies, which demonstrated that
this duration provides a sufficient perceptual window for
decision-making and aligns with previous research indicating stable
judgment levels and increased confidence beyond this exposure time
[@willis2006first].**

@@

**Participants.** Although the main part of the study relied on
within-subject design (with 109 trials per participant), we also planned
to do between-participants analyses, thus aiming at collecting a larger
sample than traditionally used in experimental psychology (with budget
availability as the main constraint). One hundred and fifty participants
were recruited via *Prolific*, a crowd-sourcing platform recognized for
providing high quality data [@peer2022; douglas2023data]. The only
inclusion criterion was a fluent proficiency in English to ensure that
the experiment instructions would be well-understood. Participants were
incentivised with a reward of about £ 7.5 for completing the study,
which took about 45 minutes to finish. Demographic variables (age,
gender, sexual orientation, education and ethnicity) were self-reported
on a voluntar

@@

The determinants of reality beliefs were modelled separately for
attractiveness, beauty, trustworthiness, and familiarity, using second
order raw polynomials coefficients to allow for possible quadratic
relationships (**Figure 2**). Aside from attractiveness (conceptualized
as a general construct), models for beauty, trustworthiness and
familiarity were adjusted for the two remaining variables *mutatis
mutandis*. The analysis focused on sexual-orientation relevant stimuli,
i.e., on faces that were aligned with respect to the participants'
sexual orientation (i.e., female faces for heterosexual males, male
faces for homosexual males, etc.), and the models included the
interaction with the participants' gender (as a sexual dimorphism has
been reported in face appraisal processes). For the attractiveness and
beauty models, we then added the interaction with the reported
self-attractiveness (the average of the two questions pertaining to it)
to investigate its potential modulatory effect. Finally, we investigated
the inter-individual correlates of simulation monitoring with similar
models (but this time, for all items regardless of the participant's
gender or sexual orientation) for each questionnaire, with all of the
subscales as orthogonal predictors.\

```{r message=FALSE, warning=FALSE, fig.cap="Figure 2. Top part shows the effect of face ratings on 1) the probability of judging a face as real vs. fake (solid line) and 2) on the confidence associated with that judgement (dashed lines) depending on the sex. Bottom part shows the effect of personality traits on the belief (black line) and the confidence associated with it (colored lines). The points are the average per participant confidence for both types of judgements. Stars indicate significance (p < .001***, p < .01**, p < .05*).", out.width="100%"}
knitr::include_graphics("../../figures/Figure2.png")
```

@@

Interestingly, we found a significant positive linear relationship in
male participants for both attractiveness and beauty on simulation
monitoring that we could interpret under an evolutionary lens.
Specifically, males purportedly place more emphasis on facial
attractiveness as a sign of reproductive potential, as compared with
females, who tend to value characteristics signaling resource
acquisition capabilities [@qi2022gender; @fink2006facial;
@buunk2002age]. It is thus possible that the evolutionary weight
associated with attractiveness skewed the perceived saliency of
attractive faces for men, rendering them significantly more salient than
unattractive faces, and in turn distorting the relationship with
simulation monitoring.
<!-- However, it is also possible that this pattern occurred due to the stimuli range, by which the stimuli did not include sufficiently unattractive faces for males to be , by which   -->
However, future studies should test this saliency-based hypothesis by
measuring constructs closer to salience and its effects, for instance
using neuroimaging [@lou2015perceptual; @indovina2007dissociation] or
physiological markers
<!-- (e.g., heart rate deceleration) [@skora2022functional]. -->[e.g.,
heart rate deceleration, @skora2022functional].

@@

Another issue is the impact on reality judgements of the prior explicit
instruction that "about half of the faces were AI-generated and the
other half real photos". Given this prior information given to
participants, it might seem like our enthusiasm pertaining to the
finding that most people did indeed believe a high number of stimuli to
be fake might be unwarranted, since it simply affirms participants
followed the instructions. However, even if that was the case, the
finding that our beliefs of reality can be so easily re-programmed with
simple instructions and lead to high-confidence answers remains an
interesting phenomenon. **Moreover, it is to note that the paradigm did
not explicitly instruct participants to balance their answers according
to a certain distribution (e.g., 50-50) - merely providing them a
description of the dataset (but participants could, and in some cases
did, deviate substantially from the information provided).** The fact
that no presentation order effect was found on reality beliefs suggests
that participants did not try to actively distribute their responses to
match the instructions, in which case we would have expected a different
pattern: for instance, the first few items judged as real (the initial
"true" belief of the participants), and a bias would progressively
appear towards responding "fake" (as participants realize that all
stimuli are of similar nature and that they have to "make up" for the
prevalence of their "real" answers to fulfil the expected proportion of
responses given the instructions).

@@

## Data Availability

@@

## Funding

This work was supported by the Presidential Postdoctoral Fellowship Grant (NTU-PPF-2020-10014) from Nanyang Technological University (awarded to DM).

@@

## References
